Modeling Intervention Strategies for United States TB Control

### 2014-07-14 in_progress/presentations/paper/draft.tex
Working on draft. This paper is definitely getting out.

\documentclass{amsart}
\usepackage[pdftex]{graphicx}
\usepackage{enumerate,placeins}
\usepackage{color}
\graphicspath{{figures/}}

@@

\section{Abstract}
An epidemiological and economic model for tuberculosis in the United States
 was developed, extending a model developed in 2012 by the Centers
for Disease Control and Prevention. Sensitivity analysis of this model reveals
that the percentage of foreign-born immigrants with LTBI is a highly influential
parameter. Various intervention strategies, particularly those that reduced LTBI
among foreign-born arrivals, were evaluated, and cost per case averted estimates
were found.  Additionally, a population-level, stochastic, agent-based model was
built. This model demonstrated the feasibility of using agent-based modeling in
the context of minimal population heterogeneity and provided statistical
characterization of the results of the compartmental model. The distributions
of the final population sizes and incidence rates projected by the stochastic model
were found to be approximately normal, with means centered at the results of
compartmental model.

\section{Introduction}
Epidemiological models allow public health professionals to predict and analyze
disease dynamics and intervention effectiveness. The most common examples of
such models are detereministic, compartmental, differential equation models. In
these models, the population is split between several possible health states, or
compartments, and a system of differential equations governs population flow
between the various compartments. In 2012, Hill, Becerra, and Castro implemented
a compartmental differential equaiton model of tuberculosis (TB) in the United
States (US).  Their model characterized five health states across two
subpopulations, US-born (USB) and foreign-born (FB), for a total of 10
compartments. They used this model to evaluate several possible intervention
strategies and ultimately concluded that though increasing LTBI treatment was an
effective intervention strategy, the US was unlikely to meet their stated goal
of elimination of TB in the US by 2100. Here, we present an extended Hill model
with additional tracking capabilities, such that it can now report
further granularity in the disease dynamics. Furthermore, this model also tracks
economic data in order to project the US health care system (HCS) costs due to
TB given our current policy as well as in the context of various interventions.

@@

\section{Background}
\subsection{Tuberculosis}
Tuberculosis is an infectious bacterial disease that primarily affects the lungs,
with common symptoms including severe coughing, fevers, and chest pain.
Tuberculosis is a major global health concern, the World Health Organization
 estimates that 8.6 million developed TB and and 1.3 million people died from
 the disease in 2012.  As an airborne disease, \emph{Mycobacterium tuberculosis}
 is typically transmitted from person-to-person by coughing, sneezing, and the 
exchange of other respiratory fluids.  Infected individuals first enter a latent phase,
 during which \emph{Mycobacterium tuberculosis} is present in their bodies,
but the individuals remain asymptomatic and noncontagious. 
The lifetime risk of progressing from Latent TB Infection (LTBI) to Active TB is approximately
5-10\%, although individuals with LTBI have the greatest chance of developing 
Active TB during the first two years immediately following infection. 
Cases of Active TB may be classified as pulmonary ($\sim$80\% of cases) or 
non-pulmonary ($\sim$20\% of cases), depending upon the biological location of infection.
In the Hill model, pulmonary Active TB is treated as contagious and non-pulmonary 
Active TB is treated as  non-contagious, which is an assumption we also use here.  \\

In the United States, 9,951 new cases of TB were reported in 2012 by the Centers for Disease
Control. The disease burden varied for different subpopulations in the U.S., and in particular
the incidence rate of TB was 11.5 times higher for the foreign-born (FB) population
than the US-born (USB) population.  For the past 20 years, the overall incidence of
TB in the U.S. has declined each year, but in 2012 the annual incidence level of 32 new cases/million
remains above the desired goal of the $<$1 new case/million, which is the target
for disease elimination set by the Centers for Disease Control. \\

citations: CDC Tuberuclosis 2013 Report - http://www.cdc.gov/mmwr/preview/mmwrhtml/mm6211a2.htm
WHO Global Tuberculosis Report 2013 - http://www.who.int/tb/publications/global\_report/en/ \\

\subsection{The Hill Model}
A flowchart representation of the Hill Model is shown in
Figure~\ref{fig:hillModelSchematic}. Each compartment represents a different
possible health state with respect to TB for every US-born or foreign-born
individual, and arrows between different compartments represent possible
transitions between states.  Individuals also leave the model from all
compartments due to natural death, which is left out of the figure for clarity.

@@

The majority of USB and FB individuals fall into the Susceptible (S) category,
which includes everyone who is uninfected and has not been exposed to TB.  After
exposure to an individual with TB, a person in the Susceptible compartment can
develop Latent TB Infection (LTBI). Latently infected individuals are
asymptomatic and non-infectious, but have some risk of developing active TB
infection over time. To reflect the fact that real LTBI patients have a
much higher risk of developing active TB within two years of exposure, the Hill
Model splits the LTBI compartment into Acute LTBI (fast progressors) and Chronic
LTBI (slow progressors). Accordingly, individuals in the Acute LTBI compartment
have a higher risk of developing active TB than those in the Chronic LTBI
compartment. Individuals in the Chronic LTBI compartment may also be
exogenously re-infected and transition to the Acute LTBI compartment.

@@

\section{Methods}
\subsection{Basic Structure}
Both the Hill Model (Basic Hill Model) and our extended Model (Extended Hill
Model) were implemented in \texttt{R} as a system of differential equations,
which were solved via the $\texttt{lsoda}$ routine. The systems of differential
equations used to capture basic disease dynamics in the Basic Hill Model are
shown in Figure~\ref{fig:hillEquations}.  In the Extended Hill Model, each flow
rate equation from the Basic Hill Model was separated into its component part
and these were tracked separately. In other words, rather than tracking net flow
into various compartments, the Extended Hill Model tracks flow between all pairs
of compartments separately. Furthermore, these equations were each also supplemented to
also estimate their corresponding US HCS costs.  These equations are detailed in
the\\
{\huge\color{red} Add extended Hill equations to the appendix.}\\
appendix,
section~\ref{sec:extendedHillEqs}.  Historical data from 2000-2008 was used to
initialize the model, which was then run up to the year 2100.

\begin{figure} 
  {\huge\color{red} Make this not a figure. This should definitely just be tex.}
  \begin{center}
    \includegraphics[scale=0.75]{figures/BasicHillEquations.pdf}
  \end{center}
  \caption{System of Differential Equations given in the Hill Model.}
  \label{fig:hillEquations}
\end{figure}

The vector variables $S_0, F_0, L_0, I_0, J_0$ contain the number of US-born
individuals in the S, F, L, I, J compartments respectively, whereas $S_1, F_1,
L_1, I_1, J_1$ contain the number of foreign-born individuals. $N_0$ and $N_1$
are the total populations of US-born and foreign-born individuals.  The
constants $\rho$ and $\alpha$ are birth rates, while $\mu_i$, and $\mu_d$ are
death rates. A complete list and descriptions of all constants used in the model
can be found in the Appendix, Section~\ref{subsec:hillConstants}.

\subsection{Additional Tracking Capabilities}
In order to refine the tracking capabilities of the Hill Model, the original
differential equations used to describe TB spread were separated into their
component parts and each section was tracked separately.
These components were made into compartments, tracked by differential equations
detailed in the appendix. Progressions into active TB due to activations of LTBI
or exogenous infection were tracked, allowing for sourced incidence data to be
generated. These equations allowed the model to track the sourced number of TB
cases, TB deaths, and natural deaths. Furthermore, the model also tracks the
sourced cost on the US HCS. The immigrating cases of LTBI (acute or chronic)
were also tracked. In the case of intervention testing, the number of cured and
untreated cases of entering LTBI were both tracked. Cured cases of LTBI
entering, TB deaths, total TB cases, and total cost were also tracked discounted
at 3\% annually. This discounting was converted to a continuous differential
equation for use within the model. In general, incidence data is calculated in
the same way in the Extended Hill Model as it was in the Basic Hill Model.

In addition, estimations of the basic reproductive number of FB or USB active,
infectious TB cases were made from a theoretical and an experimental
perspective. Experimental data were calculated by reducing the initial
FB or USB infectious TB populations by one and allowing the
model to progress otherwise as normal. The decreased number of total TB cases
seen details how many infections can be thought to be due to one infected
individual in the given population. In the agent-based model, this \emph{one}
was a true model agent whereas in the compartmental model, this decrease of
\emph{one} was a proportional decrease based on the scaled population level.

From a theoretical perspective, the spread of TB was thought to be a geometric
series. If one infectious individual infects $x$ people annually, over the
course of $N$ years the total number of infections caused by this individual can
be approximated by the geometric series of $N$ terms with rate $x$ and initial
term $1$. In this case, $N = 100$, and the ratios for FB infectious individuals
and USB infectious individuals were obtained from <+RATIO CITATION COLIN+>
\subsection{Economic Modeling}
In order to track the US HCS economic load given the TB rates, we made the
following treatment cost assumptions. A single case of active TB was assumed to
yield a \$14,000 US HCS cost, charged immediately upon disease contraction. This
cost is the weighted mean of the costs of cases requiring hospitalization
(happening 49\% of the time) and cases not requiring hospitalization (51\%).
These cost estimates were found in <+DYLAN COST STUDY CITATION+>. LTBI treatment
costs were given by \$468.00 US HCS cost, charged immediately upon successful
treatment. These estimates were calculated based on the cost of a successfull
treatment, and the typical adherence and efficacy of said treatment <+DYLAN COST
STUDY CITATION, ADHERENCE CITATION+>. It is important to note that the charging
assumptions are different for treatment of active TB as opposed to treatment of
latent TB. In particular, cases of active TB imply a cost immediately upon
disease contraction, whereas latent TB costs are charged only upon successful
treatment. The motivation for this distinction is that active TB treatment is
mandated by the US for all known cases and, further, most treatment costs are
incurred at the beginning of the treatment cycle. As such, charging for every
known case, immediately at contraction is a well motivated assumption. On the
other hand, latent TB is only rarely treated in discovered cases and the costs
are more evenly distributed. Furthermore, the choice to charge only upon
successful treatment yields a more conservative estimate of the success of any
intervention analyzed.

Cost data were separately tracked due to incoming active TB costs stemming from
activation of LTBI vs exogenous reinfection. These data were modelled as
additional compartments and also solved by $\texttt{lsoda}$.  Given the
uncertainty in measurments of LTBI treatment cost, extensive sensitivity
analysis was performed on this parameter relative to final cost outcomes. In
addition to US HCS cost, the system was implemented so as to also track
projected intervention implementation cost, given user-inputted parameters
relating to various possible intervention cost strategies. The extended Hill
Model was used to track the effect of many interventions tested by the Hill
Model. In particular, the intervention strategy of curing entering cases of LTBI
proved very promising and elimination year, final cost, and cost per case
averted were tracked for various levels of entering LTBI cure rate. The
interventions were implemented so as to take effect during the year 2013 and run
to 2100. The numerical DE solver $\texttt{lsoda}$ used was run with a time step
of $0.8$. Sensitivity analysis was performed on this parameter and reducing the
time step was shown to have minimal effect on final size or cost values.

\subsection{Agent Based}
The population level agent based model was implemented in several ways. Early
implementations were built in \texttt{Netlogo} and \texttt{Java}, but final
implementations were constructed in $\texttt{C++}$. Probabilites of agent
progression between various health states were computed given rates in the hill
model and a variety of integration approximations. The model maintained
individual records for every individual with LTBI or active TB. Susceptible
individuals were not modelled as agents; instead, a binomial distribution was used to
probabilistically determine the number of new infections each time step.
Similarlly, the number of new immigrants each time step was chosen from a
probability distribution. We collected final population distribution data via
<+NUMBER OF RUNS 2160+> runs with each agent in the model truly representing one
infected individual and a time step of $0.01$.  These data were analyzed in $R$.
This model was also made to track deaths, total TB cases, and infection sources,
as well as economic data. These data were consistently normal and matched the
deterministic model.

Further implementations of the \texttt{C++} model were made that deviated from
the basic Hill Model. Specifically, the acute latent compartment in the basic
Hill Model is a vestigal necessity of compartmental modeling and not reflective
of the biology of tuberculosis. As such, an agent based model was implemented
that more closely respects the biology of tuberculosis. Results from this model
were also normal and did not differ significantly from the strict Hill model.

\section{Results}
\subsection{Basic Population Breakdown}
The additional tracking capabilities offered several key insights into US TB
dynamics. Figure~\ref{fig:incPlotSourced} shows the yearly incidence of US TB, broken
down into infection source. It can be seen that the majority of the US TB load
is driven by activations of FB LTBI, followed by USB LTBI activations. This data
further agrees with the conclusions drawn by Hill, Becerra, and Castro about the
necessity of LTBI treatment in any effective intervention strategy. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.5]{incPlotSourced}
  \end{center}
  \caption{Sourced yearly incidence data generated by the extended Hill Model.}
  \label{fig:incPlotSourced}
\end{figure}
Further, figure~\ref{fig:costPlotSourced} shows a similarly sourced plot, but analyzing the
final US HCS costs due to active TB. One can see that in this plot, roughly half
of the US TB HCS costs are due to activations of LTBI. Note that in this data,
it is appropriate to think of the costs due to acute LTBI as costs due to
exogenous re-inection, whereas costs due to chronic LTBI are activations of long
standing latent infections. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.35]{costPlotSourced}
  \end{center}
  \caption{Sourced US HCS economic TB load. Note that this data only illustrates
    the load due to treating active TB, but illustrates where the infections
    driving this cost come from.}
  \label{fig:costPlotSourced}
\end{figure}
Note that both of these plots underestimate the impact of LTBI on the spread of
TB, as every LTBI activation contributes not only
to incidence and costs directly, but also indirectly by causing additional
future cases, which is not captured in these graphs. Further, there are also US
HCS costs due to LTBI treatment, which is not illustrated in these graphs. 
\subsubsection{Basic Reproduction Number}
The basic reproduction number of FB or USB cases of infectious TB was also
estimated by this system. Using the best-fit values of parameters from the 
Hill model, we found the average number of secondary infections arising from
a single case of infectious TB to be 0.423 and 0.370 for USB and FB populations
respectively.  These values differ because in the model, there are different
estimates for death rates and the rate of TB treatment and self-cure in the USB and FB populations, 
so the average infectious period differs between the two populations.
Extrapolating these results, we can think of the
total number of secondary infections over 100 years due to a FB or USB
infectious TB case as describing a geometric series in a large population.
Presuming there are no overlaps in infectious contacts, if a single case of
infectious TB in either population infects $p_f, p_u$ new cases,
respectively, then we can say that over the course of 100 years, the total
number of cases infected will follow a geometric series.
{\huge \color{red} TODO: formalize this with math. We can fit it here.}
This analysis predicts
that over 100 years, one USB infection will lead to $1.04$ subsequent infections,
whereas one FB infection will lead to $.83$ subsequent infections. Experimentally, 
these data were also analyzed, with results of $1.03$ and $.64$, respectively. 
Full calculations are included in the Appendix.

>>>>

% arara: pdflatex
% arara: bibtex
% arara: pdflatex
% arara: pdflatex
\documentclass{amsart}
\usepackage[pdftex]{graphicx}
\usepackage{enumerate,placeins}
\usepackage{color}
\graphicspath{{figures/}}

\newcommand{\TODO}[1]{\hfill\\{\huge \color{red} TO DO:} #1 \hfill \\}

@@

\section{Abstract}
An epidemiological and economic model for tuberculosis in the United States
 was developed, extending a model developed in 2012 by the Centers
for Disease Control and Prevention. Sensitivity analysis of this model reveals
that the percentage of foreign born immigrants with LTBI is a highly influential
parameter. Various intervention strategies, particularly those that reduced LTBI
among foreign-born arrivals, were evaluated, and cost per case averted estimates
were found.  Additionally, a population-level, stochastic, agent-based model was
built. This model demonstrated the feasibility of using agent-based modeling in
the context of minimal population heterogeneity and provided statistical
characterization of the results of the compartmental model. The distributions
of the final population sizes and incidence rates projected by the stochastic model
were found to be approximately normal, with means centered at the results of
compartmental model.

\section{Introduction}
Epidemiological models allow public health professionals to predict and analyze
disease dynamics and intervention effectiveness. The most common examples of
such models are detereministic, compartmental, differential equation models. In
these models, the population is split between several possible health states, or
compartments, and a system of differential equations governs population flow
between the various compartments. In 2012, Hill, Becerra, and Castro implemented
a compartmental differential equaiton model of tuberculosis (TB) in the United
States (US).  Their model characterized five health states across two
subpopulations, US-born (USB) and foreign-born (FB), for a total of 10
compartments. They used this model to evaluate several possible intervention
strategies and ultimately concluded that though increasing LTBI treatment was an
effective intervention strategy, the US was unlikely to meet their stated goal
of elimination of TB in the US by 2100. Here, we present an extended Hill model
with additional tracking capabilities, such that it can now report
further granularity in the disease dynamics. Furthermore, this model also tracks
economic data in order to project the US health care system (HCS) costs due to
TB given our current policy, as well as in the context of various interventions.

@@

\section{Background}
\subsection{Tuberculosis} x 
\TODO{Use TB instead of Tuberculosis maybe?}
Tuberculosis is an infectious bacterial disease that primarily affects the lungs,
with common symptoms including severe coughing, fevers, and chest pain.
Tuberculosis is a major global health concern; The World Health Organization
estimates that 8.6 million developed TB and 1.3 million people died from
the disease in 2012. Tuberculosis is an airborne disease and is typically
transmitted from person-to-person by coughing, sneezing, and the exchange of
other respiratory fluids.  Infected individuals first enter a latent phase,
during which the disease causing agent, \emph{Mycobacterium tuberculosis}, is
present in their bodies, but the individuals remain asymptomatic and
noncontagious. During this phase, individuals are said to have latent TB
infection (LTBI). Eventually, some number of individuals with LTBI progress into
what is known as the active phase of the disease, or active TB. Individuals with
an active TB infection are symptomatic and contagious. The lifetime risk of
progressing from Latent TB Infection (LTBI) to active TB is approximately
5-10\%, although individuals with LTBI have the greatest chance of developing
Active TB during the first two years immediately following infection, and if
they have a suppressed immune system, such as in the case of a comorbid HIV
infection. Cases of Active TB may be classified as pulmonary ($\sim$80\% of
cases) or non-pulmonary ($\sim$20\% of cases), depending upon the biological
location of infection. In the Hill model, pulmonary Active TB is treated as
contagious and non-pulmonary Active TB is treated as non-contagious, which is an
assumption we also use here.

In the United States, 9,951 new cases of TB were reported in 2012 by the Centers
for Disease Control. The disease burden varied for different subpopulations in
the U.S., and, in particular, the incidence rate of TB was 11.5 times higher for
the FB population than the USB population.  For the
past 20 years, the overall incidence of TB in the U.S. has declined each year,
but in 2012 the annual incidence level of 32 new cases/million remains above the
desired goal of the $<$1 new case/million, which is the target for disease
elimination set by the Centers for Disease Control.

\TODO{}citations: CDC Tuberuclosis 2013 Report - http://www.cdc.gov/mmwr/preview/mmwrhtml/mm6211a2.htm
WHO Global Tuberculosis Report 2013 - http://www.who.int/tb/publications/global\_report/en/

\subsection{The Hill Model}
A flowchart representation of the Hill Model is shown in
Figure~\ref{fig:hillModelSchematic}. Each compartment represents a different
possible TB health state for every US-born or FB individual, and
arrows between different compartments represent possible transitions between
states. Individuals also leave the model from all compartments due to natural
death, which is left out of the figure for clarity.

@@

The majority of USB and FB individuals fall into the Susceptible (S) category,
which includes everyone who is uninfected and has not been exposed to TB, or
those who have been successfully cured of the disease. After exposure to an
individual with TB, a person in the Susceptible compartment can develop Latent
TB Infection (LTBI). Latently infected individuals are asymptomatic and
non-infectious, but have some risk of developing active TB infection over time.
To reflect the fact that real LTBI patients have a much higher risk of
developing active TB within two years of exposure, the Hill Model splits the
LTBI compartment into Acute LTBI (fast progressors) and Chronic LTBI (slow
progressors). Accordingly, individuals in the Acute LTBI compartment have a
higher risk of developing active TB than those in the Chronic LTBI compartment.
Individuals in the Chronic LTBI compartment may also be exogenously re-infected
and transition to the Acute LTBI compartment.

@@

\section{Methods}
\subsection{Basic Structure}
Both the Hill Model (Basic Hill Model) and our extended Model (Extended Hill
Model) were implemented in \texttt{R} as a system of differential equations,
which were solved via the $\texttt{lsoda}$ routine. The systems of differential
equations and relevant constants used to capture basic disease dynamics in the
Basic Hill Model are shown in Figures~\ref{fig:hillEquations}
and~\ref{fig:basicHillConstants}, though a full analysis of the model is better
suited for \TODO{<+Cite Hill+>}. \TODO{Add equations for $\lambda$. I don't have
a pdf of Hill, apparently, though, so I can't.}  In the Extended Hill Model,
each flow rate equation from the Basic Hill Model was separated into the
components that tracked specific population migrations (i.e., terms that dictate
flow rates from one compartment to another) and these were tracked separately.
In other words, rather than tracking net flow into various compartments, the
Extended Hill Model tracks flow between all pairs of compartments separately.
Furthermore, these equations were each also supplemented to estimate their
corresponding US HCS costs. A relevant subset of the Extended Hill equations are
detailed in the appendix, section~\ref{sec:extendedHillEqs}. Historical data
from 2000-2008 was used to initialize the model, which was then run up to the
year 2100.

\begin{figure}
  \begin{align*}
    \dot{S}_0 &= \rho(N_0+N_1) + \sigma_0^FF_0 + \sigma^LL_0 
                +\varphi_0(I_0+J_0) - \lambda_0 S_0 - \mu_0 S_0\\
    \dot{F}_0 &= p\lambda_0S_0 + xp\lambda_0L_0 - (\mu_0+\nu^F+\sigma_0^F)F_0\\
    \dot{L}_0 &= (1-p)\lambda_0S_0 - xp\lambda_0L_0 
                -(\mu_0+\nu_0^L+\sigma^L)L_0\\
    \dot{I}_0 &= q(\nu^FF_0 + \nu_0^LL_0) - (\mu_0+\mu^d+\varphi_0)I_0\\
    \dot{J}_0 &= (1-q)(\nu^FF_0 + \nu_0^LL_0) - (\mu_0+\mu^d+\varphi_0)J_0\\
    \dot{S}_1 &= (1-f)\alpha(N_0+N_1) + \sigma_1^FF_1 + \sigma^LL_1 
                +\varphi_1(I_1+J_1) - \lambda_1S_1 - \mu_1S_1\\
    \dot{F}_1 &= gpf\alpha(N_0+N_1) + p\lambda_1S_1 + xp\lambda_1L_1 
                -(\mu_1+\nu^F+\sigma_1^F)F_1\\
    \dot{L}_1 &= (1-gp)f\alpha(N_0+N_1) + (1-p)\lambda_1S_1 - xp\lambda_1L_1 
                -(\mu_1+\nu_1^L+\sigma^L)L_1\\
    \dot{I}_1 &= q(\nu^FF_1 + \nu_1^LL_1) - (\mu_1+\mu^d+\varphi_1)I_1\\
    \dot{J}_1 &= (1-q)(\nu^FF_1 + \nu_1^LL_1) - (\mu_1+\mu^d+\varphi_1)J_1\\
  \end{align*}
  \caption{System of Differential Equations given in the Hill Model.}
  \label{fig:hillEquations}
\end{figure}
\begin{figure}
  \begin{align*}
    \sigma^{L} &= 0.057   &&\text{(Treatment rate for chronic LTBI)}\\
    v^{L}_{0}  &= 0.0014  &&\text{(Progression rate for reactivation in the USB
                                   population)}\\
    v^{L}_{1}  &= 0.0010  &&\text{(Progression rate for reactivation in the FB
                                   population)}\\
    f          &= 0.187   &&\text{(Fraction of FB arrivals with LTBI)}\\
    p          &= 0.103   &&\text{(Fraction of new infections which are acute)}\\
    ARI_{0}    &= 0.00030 &&\text{(2000 Annual Risk of Infection, USB)}\\
    q          &= 0.708   &&\text{(Fraction of infections progressing to infectious
                                   disease)}\\
    g          &= 0.0047  &&\text{(Fraction of FB arrivals with LTBI who are fast
                                   progressors)}\\
    \sigma^{F} &= 0.461   &&\text{(Cumulative fraction of treatment for acute
                                   infection)}\\
    r_{0}      &= 0.667   &&\text{(Fraction of cases due to reactivation in USB
                                   population)}\\
    r_{1}      &= 0.780   &&\text{(Fraction of cases due to reactivation in FB
                                   population)}\\
    \mu^{d}    &= 0.115   &&\text{(Mortality rate due to TB)}\\
    x          &= 0.111   &&\text{(Fraction of re-infected chronic LTBI moving to
                                   acute infection)}\\
    \phi       &= 0.897   &&\text{(Cumulative fraction self-cure and treatment of
                                   active disease)}\\
    e_{0}      &= 0.965   &&\text{(Fraction of preferred contacts within own
                                   population for USB)}\\
    e_{1}      &= 0.985   &&\text{(Fraction of preferred contacts within own
                                  population for FB)}\\
  \end{align*}
  \caption{The relevant constants used in the Basic Hill Model.  }
  \label{fig:basicHillConstants}
\end{figure}

The vector variables $S_0, F_0, L_0, I_0, J_0$ contain the number of US-born
individuals in the S, F, L, I, J compartments respectively, whereas $S_1, F_1,
L_1, I_1, J_1$ contain the number of foreign-born individuals. $N_0$ and $N_1$
are the total populations of US-born and foreign-born individuals. 
\TODO{Is the next bit needed? I think move the stuff below out and the stuff
above back where we introduce the equations.}
The constants $\rho$ and $\alpha$ are birth rates, while $\mu_i$, and $\mu_d$
are death rates. A complete list and descriptions of all constants used in the
model can be found in the Appendix, Section~\ref{subsec:hillConstants}.

\subsection{Additional Tracking Capabilities}
In order to refine the tracking capabilities of the Hill Model, the original
differential equations used to describe TB spread were separated into their
component parts and each part was tracked separately. In effect, each flow rate
between two compartments in the Basic Hill model was made into its own
``virtua'' compartment, thereby enabling computation not only of expected TB
incidence, but also of the source of that incidence. In particular, progressions
into active TB due to activations of LTBI or exogenous infection were tracked,
allowing for sourced incidence data to be generated. These equations allowed the
model to track the sourced number of TB cases, TB deaths, and natural deaths.
Furthermore, the model also tracks the sourced cost on the US HCS. The
immigrating cases of LTBI (acute or chronic) were also tracked. In the case of
intervention testing, the number of cured and untreated cases of entering LTBI
were both tracked. Cured cases of LTBI entering, TB deaths, total TB cases, and
total cost were also tracked discounted at 3\% annually. This discounting was
converted to a continuous differential equation for use within the model. In
general, incidence data is calculated in the same way in the Extended Hill Model
as it was in the Basic Hill Model.

\TODO{Where does this fit in? I'm not seeing why this is here? Maybe in the
results, instead?}
In addition, estimations of the basic reproductive number of FB or USB active,
infectious TB cases were made from a theoretical and an experimental
perspective. Experimental data were calculated by reducing the initial
FB or USB infectious TB populations by one and allowing the
model to progress otherwise as normal. The decreased number of total TB cases
seen details how many infections can be thought to be due to one infected
individual in the given population. In the agent-based model, this \emph{one}
was a true model agent whereas in the compartmental model, this decrease of
\emph{one} was a proportional decrease based on the scaled population level.

From a theoretical perspective, the spread of TB was thought to be a geometric
series. If one infectious individual infects $x$ people annually, over the
course of $N$ years the total number of infections caused by this individual can
be approximated by the geometric series of $N$ terms with rate $x$ and initial
term $1$. In this case, $N = 100$, and the ratios for FB infectious individuals
and USB infectious individuals were obtained from 
\TODO{<+RATIO CITATION COLIN+>.}
\subsection{Economic Modeling}
In order to track the US HCS TB cost, we made the following treatment cost
assumptions. A single case of active TB was assumed to yield a \$14,000 US HCS
cost, charged immediately upon disease contraction. This cost is the weighted
mean of the costs of cases requiring hospitalization (49\%) and cases not
requiring hospitalization (51\%).  These cost estimates were found in 
\TODO{<+DYLAN COST STUDY CITATION+>.} LTBI treatment costs were assumed to be
\$468.00, charged immediately upon successful treatment. These estimates were
calculated based on the cost of a successful treatment, and the typical
adherence and efficacy of said treatment 
\TODO{<+DYLAN COST STUDY CITATION, ADHERENCE CITATION+>.}
It is important to note that the charging assumptions are different for
treatment of active TB as opposed to treatment of latent TB. In particular,
cases of active TB imply a cost immediately upon disease contraction, whereas
latent TB costs are charged only upon successful treatment. The motivation for
this distinction is that active TB treatment is mandated by the US for all known
cases and, further, most treatment costs are incurred at the beginning of the
treatment cycle. As such, charging for every known case, immediately at
contraction is a well motivated assumption. On the other hand, latent TB is only
rarely treated in discovered cases and the costs are more evenly distributed.
Furthermore, the choice to charge only upon successful treatment yields a more
conservative estimate of the success of any intervention analyzed. Finally, it
is worth noting that, though accurate, these estimates of US HCS treatment cost
are not intended to be fixed parameters. The code for this model is freely
available, and as new research amends the necessary costs and new methods of
treatment are developed, these values can be modified and the model re-run to
produce a revised prediction of the US HCS TB cost.

In addition to refining the model to produce sourced US TB incidence estimates,
the additional cost modeling was also tracked in a way that preserved the source
of the US HCS cost. In particular, active TB treatment cost was tracked
separately for cases resulting from an activation of LTBI (modeled as a
progression from the Chronic Latent category) and for cases resulting from a
novel, exogenous infection (modeled as a progression from the Acute Latent
category). All cost equations were also solved by $\texttt{lsoda}$.  Given the
uncertainty in estimates of LTBI treatment cost, extensive sensitivity analysis
was performed on this parameter relative to final cost outcomes. Furthermore,
the US HCS cost caused by LTBI treatment (which was tracked) was often omitted
from Cost Benefit intervention analyses, so as to provide the most conservative
estimate of total US HCS cost given various TB interventions. In addition to US
HCS cost, the system was implemented so as to also track projected intervention
implementation cost, given user-inputted parameters relating to various possible
intervention cost strategies. The extended Hill Model was used to track the
effect of many interventions tested by the Hill Model. In particular, the
intervention strategy of curing entering cases of LTBI proved very promising and
elimination year, final cost, and cost per case averted were tracked for various
levels of entering LTBI cure rate. The interventions were implemented so as to
take effect during the year 2013 and run to 2100. The numerical DE solver
$\texttt{lsoda}$ used was run with a time step of $0.8$. Sensitivity analysis
was performed on this parameter and reducing the time step was shown to have
minimal effect on final size or cost values.

\subsection{Agent Based}
\TODO{What's the story we are trying to tell here? I mean, we are following a
  very standard format, but I think it would be better if we had a cohesive
  message.}
The population level agent based model was implemented in several ways. Early
implementations were built in \texttt{Netlogo} and \texttt{Java}, but final
implementations were constructed in \texttt{C++}. Probabilities of agent
progression between various health states were computed given rates in the hill
model and a variety of integration approximations. The model maintained
individual records for every individual with LTBI or active TB. Susceptible
individuals were not modeled as agents; instead, a binomial distribution was
used to probabilistically determine the number of new infections each time step.
Similarly, the number of new immigrants each time step was chosen from a
probability distribution. As the model was stochastic, all final data is
inherently merely a probabilistic sample; as such, many simulations are required
to draw conclusions from the model. As such, we collected final population
health statistics from approximately 2160 model simulations, with each agent in
the model truly representing one infected individual, and a time step of $0.01$.
These data were analyzed in $R$. The agent based model also tracks deaths, total
TB cases, and infection sources, and can also easily track economic data. These
data were consistently normal and matched the deterministic model. 
\TODO{Does it track economic data yet?}

Further implementations of the \texttt{C++} model were made that deviated from
the basic Hill Model. Specifically, the acute latent compartment in the basic
Hill Model is a vestigial necessity of compartmental modeling and not reflective
of the biology of tuberculosis. As such, an agent based model was implemented
that more closely respects the biology of tuberculosis. Results from this model
were also normal and did not differ significantly from the strict Hill model.

\section{Results}
\subsection{Basic Population Breakdown}
The additional tracking capabilities offered several key insights into US TB
dynamics. Figure~\ref{fig:incPlotSourced} shows the yearly incidence of US TB, broken
down into infection source. It can be seen that the majority of the US TB load
is driven by activations of FB LTBI, followed by USB LTBI activations. This data
further agrees with the conclusions drawn by Hill, Becerra, and Castro about the
necessity of LTBI treatment in any effective intervention strategy. 
\TODO{Why does it agree? We aren't doing any data analysis here, which is a
problem. We need to walk our readers through this.}
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.5]{incPlotSourced}
  \end{center}
  \caption{Sourced yearly incidence data generated by the extended Hill Model.}
  \label{fig:incPlotSourced}
\end{figure}
Further, figure~\ref{fig:costPlotSourced} shows a similarly sourced plot, but analyzing the
final US HCS costs due to active TB. One can see that in this plot, roughly half
of the US TB HCS costs are due to activations of LTBI. Note that in this data,
it is appropriate to think of the costs due to acute LTBI as costs due to
exogenous re-inection, whereas costs due to chronic LTBI are activations of long
standing latent infections. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.35]{costPlotSourced}
  \end{center}
  \caption{Sourced US HCS economic TB load. Note that this data only illustrates
    the load due to treating active TB, but illustrates where the infections
    driving this cost come from.}
  \label{fig:costPlotSourced}
\end{figure}
Note that both of these plots underestimate the impact of LTBI on the spread of
TB, as every LTBI activation contributes not only
to incidence and costs directly, but also indirectly by causing additional
future cases, which is not captured in these graphs. Further, there are also US
HCS costs due to LTBI treatment, which is not illustrated in these graphs. 
\subsubsection{Basic Reproduction Number}
The basic reproduction number of FB or USB cases of infectious TB was also
estimated by this system. Using the best-fit values of parameters from the 
Hill model, we found the average number of secondary infections arising from
a single case of infectious TB to be 0.423 and 0.370 for USB and FB populations
respectively.  These values differ because in the model, there are different
estimates for death rates and the rate of TB treatment and self-cure in the USB and FB populations, 
so the average infectious period differs between the two populations.
Extrapolating these results, we can think of the
total number of secondary infections over 100 years due to a FB or USB
infectious TB case as describing a geometric series in a large population.
Presuming there are no overlaps in infectious contacts, if a single case of
infectious TB in either population infects $p_f, p_u$ new cases,
respectively, then we can say that over the course of 100 years, the total
number of cases infected will follow a geometric series.
{\huge \color{red} TODO: formalize this with math. We can fit it here.}
This analysis predicts
that over 100 years, one USB infection will lead to $1.04$ subsequent infections,
whereas one FB infection will lead to $.83$ subsequent infections. Experimentally, 
these data were also analyzed, with results of $1.03$ and $.64$, respectively. 
Full calculations are included in the Appendix.
